Information carried by thalamocortical axons enables self-organization of the barrel field

### 2019-11-19 paper/briefpaper/brief.tex
Reformatting paragrapohs with a set-fill-column value of 78

\title{Self-organization of whisker barrels in simulation requires competition between thalamocortical axons}
% While writing, don't stop for errors
%\nonstopmode

@@

\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: seb.james\@sheffield.ac.uk}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Keyword 1 $|$ Keyword 2 $|$ Keyword 3 $|$ ...}

@@

\dropcap{S}patial patterns in neural connectivity provide clues about
the constraints under which brains evolve and develop. Whether or not
patterning governs information processing, models of the development
of information processing circuits can be validated in terms of their
ability to account for biological patterning
(\citealp{purves_iterated_1992}; see also
\citealp{wilson_what_2015,bednar_cortical_2016}). One of the most
distinctive patterns of neural connectivity is found in the rodent
barrel cortex
(\citealp{woolsey_pattern_1948,woolsey_structural_1970,welker_structure_1974},
see \citealp{fox_barrel_2008}). Understanding how this particular
pattern forms might shed light on the interplay of genetic and
extrinsic factors that shape cortical circuits, and on a wealth of
anatomical, physiological, and behavioural data obtained using this
model system.

The barrels are large cortical columns, identifiable from birth in layer 4 of
rodent primary somatosensory cortex as dense clusters of thalamocortical
axons, which are enclosed by borders a few neurons thick from postnatal day 3
\citep{erzurumlu_development_2012}. In the plane tangential to the cortical surface the
barrels constitute a map of the arrangement of the whiskers on the
face (see \citealp{yamakado_subdivision_1979}), such
that the cells of adjacent barrels respond most strongly and most quickly to
deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. The barrels
reflect subcortical whisker maps comprising cellular aggregates known as
barrelettes in the brainstem and barreloids in the thalamus
\citep{killackey_pattern_1980}.

@@

With the addition of the competition term, $\chi$, it is necessary to
introduce a normalisation step, in order to prevent one TC type from
{out-competing every other TC type. Normalisation represents the idea
that for a fixed number of afferent thalamocortical axons, the overall
number of branches must have some upper limit; given the finite
resources available in each axon for building new processes. We
divisively normalize the branching density, dividing it at each
computational step by the mean branching density per unit area. It is
then multiplied by the branching density per unit area at time 0, to
account for the possibility that there might be more axons afferent
for TC type $1$ cf.~TC type $2$:
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma}
\end{equation}
%
The prime symbols indicate the use of the intermediate value of $a_i$
after it has been computed according to Eqs.~\ref{eq:dc}--\ref{eq:branch_comp}
on the basis of the system state at time $t-{\delta}t$.

>>>>

%\title{Self-organization of whisker barrels in simulation requires competition between thalamocortical axons}
%\title{Simulated barrel cortex self-organizes using information carried by thalamocortical axons}
\title{Information carried by thalamocortical axons enables self-organization of the barrel field}
% While writing, don't stop for errors
%\nonstopmode

@@

\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: \href{mailto:seb.james@sheffield.ac.uk}{\nolinkurl{seb.james@sheffield.ac.uk}}}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Barrel cortex $|$ Self-organization $|$ Somatotopy $|$ Center-point $|$ Pattern $|$}

@@

\dropcap{S}patial patterns in neural connectivity provide clues about the
constraints under which brains evolve and develop. Whether or not patterning
governs information processing, models of the development of information
processing circuits can be validated in terms of their ability to account for
biological patterning (\citealp{purves_iterated_1992}
%; see also \citealp{wilson_what_2015,bednar_cortical_2016}
). One of the most
distinctive patterns of neural connectivity is found in the rodent
barrel cortex \cite{woolsey_structural_1970}.
%
%(\citealp{woolsey_pattern_1948,woolsey_structural_1970,welker_structure_1974},
%see \citealp{fox_barrel_2008}).
%
Understanding how this particular
pattern forms might shed light on the interplay of genetic and
extrinsic factors that shape cortical circuits.
%, and on a wealth of
%anatomical, physiological, and behavioural data obtained using this
%model system.

The barrels are large cortical columns, identifiable from birth in layer 4 of
rodent primary somatosensory cortex as dense clusters of thalamocortical
axons, which are enclosed by borders a few neurons thick from postnatal day 3
\citep{erzurumlu_development_2012}. In the plane tangential to the cortical
surface the barrels constitute a map of the arrangement of the whiskers on the
face
%(see \citealp{yamakado_subdivision_1979})
, such that the cells of adjacent barrels respond most strongly and most
quickly to deflection of adjacent whiskers
\citep{armstrong-james_flow_1992}. The barrels reflect subcortical whisker
maps comprising cellular aggregates known as barreloids in the thalamus and
barrelettes in the brainstem.
%\citep{killackey_pattern_1980}.

@@

With the addition of the competition term, $\chi$, it is necessary to
introduce a normalisation step, in order to prevent one TC type from
out-competing every other TC type. Normalisation represents the idea that for
a fixed number of afferent thalamocortical axons, the overall number of
branches must have some upper limit; given the finite resources available in
each axon for building new processes. We divisively normalize the branching
density, dividing it at each computational step by the mean branching density
per unit area. It is then multiplied by the branching density per unit area at
time 0, to account for the possibility that there might be more axons afferent
for TC type $1$ cf.~TC type $2$:
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma}
\end{equation}
%
The prime symbols indicate the use of the intermediate value of $a_i$ after it
has been computed according to Eqs.~\ref{eq:dc}--\ref{eq:branch_comp} on the
basis of the system state at time $t-{\delta}t$.
